{\Huge Supersymmetry} \\ \vspace{5mm} Lecture notes for FYS5190/FYS9190 Supersymmetry

### 2021-11-14 pheno.tex
Update of section 7.1

%%%%%%%%%%%%%%%%%%
% Standalone document
\documentclass[notes.tex]{subfiles}
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Sparticle phenomenology}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this chapter we discuss the phenomenology of supersymmetric models and how to search for supersymmetry in experiments. We begin by returning to supersymmetry breaking in order to define some reasonable and (partially) motivated subsets of the 124 MSSM parameters which can be used to define more constrained models. We then discuss supersymmetry at hadron and lepton colliders, and finally look at precision measurements that are indirectly sensitive to the existence of sparticles.

%%%%%%%%%%%%%%%%%%%%%%%
\section{Models for supersymmetry breaking} 
%%%%%%%%%%%%%%%%%%%%%%%
Let us take a little closer look at the models we use to motivate supersymmetry breaking, $\cancel{\text{SUSY}}$-models, and what their phenomenological consequences are. This is important to keep in mind as most searches for supersymmetry are interpreted under certain assumptions on the $\cancel{\text{SUSY}}$-mechanism.

Generically such models can be illustrated as shown in Fig.~\ref{SSB}. There is one or more {\bf hidden sector} (HS) scalar superfield $X$ --- by hidden we mean that it has no or very small direct couplings to the MSSM fields --- that has an {\it effective (non-renormalizable) coupling} to the MSSM scalar fields $\Phi_i$ of the form
\begin{equation}
\mathcal{L}_{HS} = -\frac{1}{M}(\bar\theta\bar\theta)X\Phi_i\Phi_j\Phi_k,
\label{eq:HS_MSSM}
\end{equation}
where $M$ is some large scale, {\it e.g.}\ the Planck scale, that suppresses the interaction. Figure \ref{SUSYB} shows an interaction that can lead to such terms, where $M$ is the mass scale of some mediating particle $Y$. If the hidden sector is constructed so that $X$ develops a vev for its auxillary $F$-component field, $F_X$,\begin{equation}
\langle X \rangle = \theta \theta \langle F_X \rangle,
\end{equation}
it breaks supersymmetry, see the discussion of Eq.~(\ref{eq:Fbreaking}). Then (\ref{eq:HS_MSSM}) will produce a soft-term of the form of the second term in Eq.~(\ref{eq:gensoftterms}),
\begin{equation}
\mathcal{L}_{\rm soft}=-\frac{\langle F_X \rangle}{M}A_iA_jA_k,
\end{equation}
with the soft mass
\[m_{\rm soft}\ = \frac{\langle F_X \rangle}{M}.\]
This has reasonable limits in that $m_{\rm soft} \to 0$ as $\langle F_X \rangle \to 0$, which is the limit of no $\cancel{\text{SUSY}}$, and $m_{\rm soft} \to 0$ as $M \to \infty$, where the scale of the HS interaction is decoupled (the mediating particle $Y$ becomes too heavy to have any influence). We will now look at two possible ways to construct such a hidden sector called Planck-scale Mediated Supersymmetry Breaking (PMSB) and Gauge Mediated Supersymmetry Breaking (GMSB).

\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/structure.eps} 
\caption{A generic illustration of how to generate soft breaking terms~\cite{Martin:1997ns}. \label{SSB}}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[scale=0.9]{figures/SUSYB.eps} 
\caption{Interactions leading to effective 4-particle couplings in our example. \label{SUSYB}}
\end{center}
\end{figure}

%%%
\subsection{Planck-scale Mediated Supersymmetry Breaking (PMSB)}
%%%
In Planck-scale mediated $\cancel{\text{SUSY}}$ (PMSB) we blame some gravity mechanism for mediating the $\cancel{\text{SUSY}}$ from the hidden sector to the MSSM so that the scale of the breaking is $M = M_P = 2.4 \cdot 10^{18}$\,GeV. Then we need to have\footnote{The use of $\sqrt{\langle F\rangle}$ is just a conventional shorthand notation for the magnitude of the vev of whichever $F$-term that breaks supersymmetry. This is called the supersymmetry breaking scale.} $\sqrt{\langle F\rangle} \sim 10^{10}-10^{11}$\,GeV in order to get $m_{\rm soft} \simeq 100-1000$\,GeV, which is of the right magnitude not to re-introduce the hierarchy problem. The complete soft terms can then be shown to be
\begin{eqnarray}
\mathcal{L}_{\rm soft} &=& -\frac{\langle F_X \rangle}{M_P}\left(\frac{1}{2} f_a \lambda^a \lambda^a + \frac{1}{6} y'_{ijk}A_i A_j A_k + \frac{1}{2}\mu'_{ij}A_iA_j + \frac{\langle F_X \rangle^*}{M_P^2}x_{ijk}A_i^*A_jA_k + {\rm c.c.}\right)\nonumber\\
 &&- \frac{|\langle F_X \rangle |^2}{M_P^2}k_{ij}A_iA_j^*.
 \end{eqnarray}
Incidentally, we can now see why we assumed the maybe-soft breaking terms to be unimportant, as  in this model they are suppressed by $\langle F_X \rangle^*/M_P^2$ compared to the other masses. If one assumes a minimal form for the parameters at the GUT scale, motivated by the wish for unification, {\it i.e.}\ $f=f_a$, $y'_{ijk} = \alpha y_{ijk}$, $\mu'_{ij} = \beta \mu, k_{ij} = k\delta_{ij}$ then all the soft terms are fixed by just four parameters
\[m_{1/2} = f\frac{\langle F_X \rangle}{M_P}, \indent m_0^2 = k \frac{|\langle F_X \rangle|^2}{M_P^2}, \indent A_0 = \alpha \frac{\langle F_X \rangle }{M_P}, \indent B_0 = \beta\frac{\langle F_X \rangle}{M_P}.\]
The resulting phenomenology is called {\bf minimal supergravity}, {\bf mSUGRA/CMSSM}, minimal in the sense of the form of the parameters, and is the most studied, but perhaps not best motivated, version of the MSSM. Often $B_0$ and $|\mu|$ are exchanged for $\tan\beta$ at low scales using the EWSB condition  in Eq.~(\ref{eq:EWSB_condition}), so it is common to say that there are four and a half parameters in the model: $m_{1/2}$, $m_0$, $A_0$, $\tan\beta$ and ${\rm sgn}\,\mu$.

%%%
\subsection{Gauge Mediated Supersymmetry Breaking (GMSB)} 
%%%
An alternative to PMSB is gauge-mediated $\cancel{\text{SUSY}}$ where soft terms come from loop diagrams with {\bf messenger} superfields that get their own mass by coupling to the HS $\cancel{\text{SUSY}}$ vev, and that have SM gauge interactions. By dimensional analysis we must have
\[m_{\rm soft} = \frac{\alpha_i}{4\pi}\frac{\langle F \rangle}{M_{\rm messenger}}.\]
If now $\sqrt{\langle F\rangle}$ and $M_{\rm messenger}$ are roughly comparable in size then $\sqrt{\langle F \rangle} \simeq 10$\,TeV can give a viable sparticle spectrum. Notice that there is now a lot less RGE running for the parameters since the soft masses are given at a rather low scale.

One way of thinking about how these mass terms appear is that the messenger field(s) get masses from HS vevs and contribute to {\it e.g.}\ gaugino mass terms through diagrams such as the one in Fig.~\ref{GMSB}, where messenger scalars and fermions run in the loop. Note that scalars can only get mass contributions like this at two-loop order. To keep GUT unification messengers are often assumed to have small mass splittings and come in $N_5$ complete $\mathbf{5} + \overline{\mathbf{5}}$ representations of $SU(5)$.

\begin{figure}[h!]
\begin{center}
\includegraphics[scale=0.8]{figures/GMSB.eps} 
\caption{Diagram for GMSB. The messenger scalars and fermions run in the loop.\label{GMSB}}
\end{center}
\end{figure}

The minimal parametrization of GMSB models is in terms of $\Lambda = \frac{\langle F\rangle}{M_{\rm messenger}}$, $M_{\rm messenger}$, $N_5$ and $\tan\beta$ for the EWSB criterion (instead of $\mu$). This gives the soft masses
\begin{eqnarray}
M_i &=& \frac{\alpha_i}{4\pi}\Lambda N_5,\label{eq:GMSBgauginomass}\\
m_j^2 &=& 2\Lambda^2 N_5 \sum C(A)_i\left(\frac{\alpha_i}{4\pi}\right)^2.
\end{eqnarray}
While this looks independent of $M_{\rm messenger}$, the messenger scale sets the starting point of the RGE running of the sparticle masses, and thus influences their magnitude. One should notice that this gives the same hierarchy of gaugino masses as in mSUGRA, $M_3>M_2>M_1$, since (\ref{eq:GMSBgauginomass}) is ordered in terms of the strength of the gauge couplings $\alpha_i$. The origin of the hierarchy is different since in mSUGRA it comes from the running of the parameters down from the GUT scale.

>>>>

%%%%%%%%%%%%%%%%%%
% Standalone document
\documentclass[notes.tex]{subfiles}
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Sparticle phenomenology}
\label{chap:pheno}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this chapter we discuss the phenomenology of supersymmetric models and how to search for low-energy particle realisations of supersymmetry in experiments. We begin by returning to supersymmetry breaking in order to define some reasonable and (partially) motivated subsets of the 124 MSSM parameters which can be used to define more constrained models. We then discuss supersymmetry at lepton and hadron colliders, and finally look at precision measurements that are indirectly sensitive to the existence of sparticles.

%%%%%%%%%%%%%%%%%%%%%%%
\section{Models for supersymmetry breaking} 
%%%%%%%%%%%%%%%%%%%%%%%
Let us begin by taking a closer look at the models we use to motivate supersymmetry breaking, and what their phenomenological consequences are. This is important to keep in mind as most searches for supersymmetry are interpreted under certain assumptions on the supersymmetry  breaking mechanism.

Generically such models can be illustrated as shown in Fig.~\ref{SSB}. There is one or more {\bf hidden sector} (HS) scalar superfield $X$ -- by hidden we mean that it has no or very small direct couplings to the MSSM fields -- that has an {\it effective (non-renormalizable) coupling} to the MSSM scalar fields $\Phi_i$ of the form
\begin{equation}
\mathcal{L}_\text{HS} = -\frac{1}{M}(\bar\theta\bar\theta)X\Phi_i\Phi_j\Phi_k,
\label{eq:HS_MSSM}
\end{equation}
where $M$ is some large scale, {\it e.g.}\ the Planck scale, that suppresses the interaction. Figure \ref{SUSYB} shows interactions that can lead to such terms, where $M$ is the mass scale of some {\bf mediator} particle $Y$.

\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/structure} 
\caption{A generic illustration of how to generate soft breaking terms~\cite{Martin:1997ns}. \label{SSB}}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[scale=0.9]{figures/SUSYB} 
\caption{Interactions leading to effective 4-particle couplings in our example. \label{SUSYB}}
\end{center}
\end{figure}

If the hidden sector is constructed so that $X$ develops a vev for its auxillary $F$-component field, $F_X$,\begin{equation}
\langle X \rangle = \theta \theta \langle F_X \rangle,
\end{equation}
it breaks supersymmetry spontaneously, see the discussion leading up to Eq.~(\ref{eq:Fbreaking}). As a result, the interaction in  (\ref{eq:HS_MSSM}) will produce a soft-term of the form of the second term in Eq.~(\ref{eq:soft_terms_component_fields}),
\begin{equation}
\mathcal{L}_\text{soft}=-\frac{\langle F_X \rangle}{M}A_iA_jA_k,
\end{equation}
with the soft-mass parameter
\[m_\text{soft} = \frac{\langle F_X \rangle}{M}.\]
This has reasonable limits in that $m_\text{soft}  \to 0$ as $\langle F_X \rangle \to 0$, which is the limit of no supersymmetry breaking, and $m_{\rm soft} \to 0$ as $M \to \infty$, where the interaction with the hidden sector is decoupled because the mediating particle $Y$ becomes too heavy to have any influence.

We will now look at two possible ways to construct such a hidden sector called Planck-scale Mediated Supersymmetry Breaking (PMSB) and Gauge Mediated Supersymmetry Breaking (GMSB).

%%%
\subsection{Planck-scale Mediated Supersymmetry Breaking (PMSB)}
%%%
In Planck-scale mediated supersymmetry breaking (PMSB) we blame some gravity mechanism for mediating the supersymmetry breaking from the hidden sector to the MSSM so that the scale of the breaking is $M = M_P = 2.4 \cdot 10^{18}$\,GeV. Then we need to have $\sqrt{\langle F_X\rangle} \sim 10^{10}-10^{11}$\,GeV in order to get $m_{\rm soft} \simeq 50-5000$\,GeV, which is roughly of the right magnitude not to re-introduce the hierarchy problem. The use of $\sqrt{\langle F_X\rangle}$ is just a conventional shorthand notation for the magnitude of the vev of whichever $F$-term that breaks supersymmetry. This is called the {\bf supersymmetry breaking scale}.

The complete soft terms for such a mechanism can then be shown to be
\begin{eqnarray}
\mathcal{L}_\text{soft} &=& -\frac{\langle F_X \rangle}{M_P}\left(\frac{1}{2} f_i \lambda^a_i \lambda^a_i + \frac{1}{6} y'_{ijk}A_i A_j A_k + \frac{1}{2}\mu'_{ij}A_iA_j + \frac{\langle F_X \rangle^*}{M_P^2}x_{ijk}A_i^*A_jA_k + {\rm c.c.}\right)\nonumber\\
 &&- \frac{|\langle F_X \rangle |^2}{M_P^2}k_{ij}A_iA_j^*.
 \end{eqnarray}
Incidentally, we can now see why we assumed the maybe-soft breaking terms to be unimportant, as  in this model they are suppressed by $|\langle F_X \rangle|/M_P^2$ compared to the other masses.

If one assumes a minimal form for the parameters at the GUT scale, motivated by the wish for unification, {\it e.g.}\ $f=f_i$, $y'_{ijk} = \alpha y_{ijk}$, $\mu'_{ij} = \beta \mu$, and $k_{ij} = k\delta_{ij}$, then all the soft terms are fixed by just four parameters
\[m_{1/2} = f\frac{\langle F_X \rangle}{M_P}, \indent m_0^2 = k \frac{|\langle F_X \rangle|^2}{M_P^2}, \indent A_0 = \alpha \frac{\langle F_X \rangle }{M_P}, \indent B_0 = \beta\frac{\langle F_X \rangle}{M_P}.\]
The resulting phenomenology is called {\bf minimal supergravity}, or {\bf mSUGRA/CMSSM}. This is minimal in the sense of the form of the parameters, and is the most studied, but perhaps not best motivated, version of the MSSM. Usually, $B_0$ and $|\mu|$ are exchanged for $\tan\beta$ at low scales using the EWSB condition  in Eq.~(\ref{eq:EWSB_condition}), so it is common to say that there are four and a half parameters in the model: $m_{1/2}$, $m_0$, $A_0$, $\tan\beta$ and ${\rm sgn}\,\mu$.

%%%
\subsection{Gauge Mediated Supersymmetry Breaking (GMSB)} 
%%%
An alternative to PMSB is gauge-mediated supersymmetry breaking where soft terms come from {\it loop diagrams} with {\bf messenger} superfields that get their own mass by coupling to the hidden sector supersymmetry breaking  vev, and that have Standard Model gauge interactions. By dimensional analysis we must have
\[m_{\rm soft} = \frac{\alpha_i}{4\pi}\frac{\langle F \rangle}{M}.\]
If now the supersymmetry breaking scale  $\sqrt{\langle F\rangle}$ and the messenger mass $M$ are roughly comparable in size, which is reasonable given where the messenger mass comes from, then $\sqrt{\langle F \rangle} \simeq 100$\,TeV can give a viable sparticle spectrum. Notice that there is now a lot less RGE running for the parameters in the GMSB compared to the PMSB since the soft masses are created at a rather low scale.

One way of thinking about how these mass terms appear is that the messenger field(s) get masses from hidden sector vevs and contribute to for example gaugino mass terms through diagrams such as the one in Fig.~\ref{fig:GMSB}, where messenger scalars and fermions run in the loop, and their masses from the hidden sector vevs are symbolised by the mass insertions. Note that scalars can only get mass contributions like this at two-loop order since the messenger interaction is a gauge interaction, involving gauge bosons or gauginos in the MSSM. In order not to spoil GUT unification messengers are often assumed to have small mass splittings and come in $N_5$ complete $\mathbf{5} + \overline{\mathbf{5}}$ (fundamental) representations of $SU(5)$.

\begin{figure}[h!]
\begin{center}
\includegraphics[scale=0.8]{figures/GMSB} 
\caption{Diagram for GMSB giving masses to the gauginos. The messenger scalars and fermions run in the loop.\label{fig:GMSB}}
\end{center}
\end{figure}

The minimal parametrisation of GMSB models is in terms of the scale $\Lambda = \frac{\langle F\rangle}{M}$, the messenger mass $M$, the number of representations $N_5$, and $\tan\beta$ for the EWSB criterion (instead of $\mu$ and $b$). This gives the soft masses
\begin{eqnarray}
M_i &=& \frac{\alpha_i}{4\pi}\Lambda N_5,\quad\text{(gaugino masses)}\label{eq:GMSBgauginomass}\\
m_j^2 &=& 2\Lambda^2 N_5 \sum C(R)_i\left(\frac{\alpha_i}{4\pi}\right)^2,\quad\text{(scalar masses)}
\end{eqnarray}
where the sum is over the quadratic Casimir invariant for the scalar superfield $\Phi_j$ that the scalar field belongs to. We clearly see that the scalar soft-masses are a two-loop effect as discussed above.

While this parameterisation looks independent of $M$, the messenger scale sets the starting point of the RGE running of the sparticle masses, and thus influences their magnitude. For example, the tri-linear soft-term couplings $a_{ijk}$ are expected to be very small at the messenger scale, and are effectively set to zero, however, due to RGE running they are small, but non-zero at the electroweak scale. Since the scalar masses $m_j$ scale as $\sqrt{N_5}$ compared to $N_5$ for the gauginos, we in general expect the scalars to be lighter in GMSB models. One should also notice that this parameterisation  gives the same hierarchy of gaugino masses as in mSUGRA, $M_3>M_2>M_1$, since (\ref{eq:GMSBgauginomass}) is ordered in terms of the strength of the gauge couplings $\alpha_i$. The origin of the hierarchy is different since in mSUGRA it comes from the running of the parameters down from the GUT scale.
